---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r libraries, message=FALSE, warning=FALSE}
# Load r libraries that we use
library(dplyr)
library(tibble)
library(data.table)
library(ggplot2)
library(knitr)
library(kableExtra)
library(impute)
library(lubridate)

```


```{r load and preprocess, warning=FALSE}
# Check if file exists and download it and unzip it if it does not
if (!file.exists("zipped_data.zip")){
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
                  "zipped_data.zip")
    unzip("zipped_data.zip")
}

# load the data
activity <- as_tibble(fread("activity.csv"))
```


## What is mean total number of steps taken per day?

**1. Make a histogram of the total number of steps taken each day**
```{r total steps hist}

# get total steps 
tsteps <- activity %>%
    group_by(date) %>%
    summarize(total_steps = sum(steps, na.rm = TRUE))

# histogram of total steps
hg <- ggplot(data = tsteps, mapping = aes(total_steps))
hg + 
    geom_histogram(binwidth = 1000) + 
    theme_bw() +
    labs(title = "Histogram of total steps taken in a day",
         x = "Total Steps",
         y = "Frequency (days)")
```

**2. Calculate and report the mean and median total number of steps taken per
day**
``` {r mean/median steps}
# table of mean and median total steps in a day
kable(
    tsteps %>%
        summarize(mean_daily_steps = round(mean(total_steps, na.rm=T), 2),
                  median_daily_steps = round(median(total_steps, na.rm = T), 2))
) %>%
    kable_styling(full_width = F)

```



## What is the average daily activity pattern?

**1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
and the average number of steps taken, averaged across all days (y-axis)**
```{r plot daily activity}
steps_interval <- activity %>%
    group_by(interval) %>%
    summarize(avg_daily_steps = mean(steps, na.rm = T))

sg <- ggplot(data = steps_interval, mapping = aes(interval, avg_daily_steps))

sg +
    geom_line()+ 
    theme_bw() +
    labs(title = "Average steps taken in each 5-minute interval",
         x = "Interval",
         y = "Avg. Daily Steps")
```

**2. Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?**

Interval **`r steps_interval[[which.max(steps_interval$avg_daily_steps),1]] `**
contains the maximum number of steps, max is **`r round(steps_interval[[which.max(steps_interval$avg_daily_steps),2]],2)`** steps.



## Imputing missing values

**1. Calculate and report the total number of missing values in the dataset
(i.e. the total number of rows with NAs)** 

There are **`r sum(is.na(activity$steps))`** / **`r length(activity$steps)`**
missing values, or **`r round(100*mean(is.na(activity$steps)), 1)`**%.
(steps are the only value that is missing in the dataset)

**2. Devise a strategy for filling in all of the missing values in the dataset. The
strategy does not need to be sophisticated. For example, you could use
the mean/median for that day, or the mean for that 5-minute interval, etc.**  

```{r examine where missing values are}
# Examine where we see missing values

# count of intervals, grouped by percentage of missing values for that interval
activity %>%
    group_by(interval) %>%
    summarize(total = n(),
              missing = sum(is.na(steps))) %>%
    mutate(pct_missing = round(100*missing/total,1)) %>%
    group_by(pct_missing) %>%
    summarize(number_of_intervals = n())

# count of days grouped by percentage of missing values on that day
activity %>%
    group_by(date) %>%
    summarize(total = n(),
              missing = sum(is.na(steps))) %>%
    mutate(pct_missing = round(100*missing/total,1)) %>%
    group_by(pct_missing) %>%
    summarize(number_of_days = n())
```
*It would appear that days that have any missing values are fully empty. Imputing
missing values by mean/median on a day would not work based on that. We will 
instead impute by median for that interval*

``` {r prep medians for imputation}
# Prepare medians for imputation
interval_medians <- activity %>%
    group_by(interval) %>%
    summarize(int_med = median(steps, na.rm = T))
```
*(Consideration: this is somewhat naive and zero-skewed as the median is 
oftentimes zero for an interval, but mean will always be above zero which is 
not practical either.)*


**3. Create a new dataset that is equal to the original dataset but with the
missing data filled in.**  

```{r patch missing values}

activity_nomissing <- activity %>%
    inner_join(interval_medians) %>%
    mutate(steps = ifelse(is.na(steps), int_med, steps)) %>%
    select(date, interval, steps)

```


**4. Make a histogram of the total number of steps taken each day and Calculate
and report the mean and median total number of steps taken per day.**
  

```{r histo plot - round 2}

tsteps_nomissing <- activity_nomissing %>%
    group_by(date) %>%
    summarize(total_steps = sum(steps))

ag <- ggplot(tsteps_nomissing, mapping = aes(total_steps))

ag + 
    geom_histogram(binwidth = 1000) + 
    theme_bw() + 
    labs(title = "Histogram of Total Daily Steps",
         x = "Total Steps",
         y = "Number of Days")
```
``` {r mean and median table - round 2}
kable(
    tsteps_nomissing %>%
        summarize(mean_daily_steps = round(mean(total_steps, na.rm=T), 2),
                  median_daily_steps = round(median(total_steps, na.rm = T), 2))
) %>%
    kable_styling(full_width = F)


```

*Do
these values differ from the estimates from the first part of the assignment?*

We see that the median has been unchanged (since all values were imputed from 
medians) but that the mean has increased (since missing values were removed and
therefore treated as zero). 

*What is the impact of imputing missing data on the estimates of the total
daily number of steps?*

The impact on imputation is that total steps will increase since NA's are 
removed in prior calculations and now are being included.




## Are there differences in activity patterns between weekdays and weekends?

*For this part the weekdays() function may be of some help here. Use the dataset
with the filled-in missing values for this part.*


**1. Create a new factor variable in the dataset with two levels – “weekday”
and “weekend” indicating whether a given date is a weekday or weekend
day.**

```{r create weekpart factor}


activity_nomissing <- activity_nomissing %>%
    mutate(wkday = weekdays(date)) %>%
    mutate(weekpart = as.factor(
        ifelse(wkday == "Saturday" | wkday == "Sunday",
               "Weekend",
               "Weekday"))) %>%
    select(weekpart, date, interval, steps)
```


**2. Make a panel plot containing a time series plot (i.e. type = "l") of the
5-minute interval (x-axis) and the average number of steps taken, averaged
across all weekday days or weekend days (y-axis).**

```{r weekpart agg, message=FALSE}
# Aggregate data by weekpart (weekday / weekend)

wkpt_agg <- activity_nomissing %>%
    group_by(weekpart, interval) %>%
    summarize(avg_steps = mean(steps))

```



```{r plot weekday vs. weekend avg steps}
# Plot the data
wg <- ggplot(wkpt_agg, mapping = aes(interval, avg_steps))

wg + 
    geom_line() + 
    theme_bw() + 
    facet_grid(rows = vars(weekpart)) + 
    labs(title = "Average Steps taken in each 5-minute interval",
         x = "Interval",
         y = "Number of Steps")

```

There are two noticeable patterns that can be observed when we compare weekday 
and weekend activity:

1. Activity seems to start and end later in the day on weekends vs weekdays
2. There seems to be slightly higher average steps throughout the day on 
weekends vs. weekdays